---
title: "Assin 1"
author: "Brian Cox"
date: "2023-01-11"
output: html_document
---

## TAKE HOME ASSIGNMENT
---
DUE: Wednesday January 19th 11:59 pm

For this assignment you will apply what you learned about basic
statistical analysis to a data set on heart disease. The assignment
should be completed as a new RMD document and knitted. Submit the html
knitted document and the original RMD document.

*Assignment* To help you out I have set up another data set.

To get you started here is the link to the data
<https://archive.ics.uci.edu/ml/machine-learning-databases/statlog/heart/heart.dat>

and here is a vector of column names:

age, sex, chestPain, restBP, cholesterol, fastGlucose, restEC,
maxHeartRate, exAngina, oldPeak, slope, numVessels, thal, disease

Here is what they mean

      age-- 1. age       
      sex-- 2. sex       
      cp-- 3. chest pain type  (4 values)       
      restbp-- 4. resting blood pressure  
      chol-- 5. serum cholesterol in mg/dl      
      fbs-- 6. fasting blood sugar > 120 mg/dl       
      restecg-- 7. resting electrocardiographic results  (values 0,1,2) 
     maxach -- 8. maximum heart rate achieved  
     exang -- 9. exercise induced angina    
     oldpeak -- 10. oldpeak = ST depression induced by exercise relative to rest   
    slope  -- 11. the slope of the peak exercise ST segment     
    num  -- 12. number of major vessels (0-3) colored by fluoroscopy        
    thal -- 13.  thal: 3 = normal; 6 = fixed defect; 7 = reversible defect
    disease -- 14. 1 = healthy; 2 = sick

*Attributes types*

Real: 1,4,5,8,10,12 Ordered:11, Binary: 2,6,9 Nominal:7,3,13

*QUESTIONS*

Starting with a new notebook,completed the numbered tasks/questions
below. Put each task as at least one code chunk or more if needed.

## Concept quesitons

Complete this question on your own using short answers 1-2 sentences.

1.  What is the difference between the functions c() and paste()?


2.  What is one similarity and one major difference between a matrix() and a data.frame() object?


3.  Explain what each of these lines of code will do?

myVar <- patient_ID 

myVar2 <- "123663"

myVar3 <- list(A, 23, "X")


## Coding activity

*Be sure to briefly describe what each of your code chunks is doing and
why you are doing the analysis/calculation.*

Only use the # comments in the code chunks to organize and annotate the
code, DO NOT use this to discuss results.

After the code chunk explain your findings or interpret you results the free text. For example if using a statistical test was the result significant ornon-significant? Next make any conclusions from your
findings. You can work alone or in a small group of 2-3 people. But you
must submit your own work.

### 1. Read in the data tables. We will do this together!

This code chunk will read in the data remotely from a url and add the
labels on to the column and format some of the categorical variables to
be more readable. Often these are 0 and 1 or 1 and 2. It is convent to
people readable letters or words. For example sex will be converted from
0 and 1 to M and F.

```{r}
# url<-"https://archive.ics.uci.edu/ml/machine-learning-databases/statlog/heart/heart.dat"
# 
# heart <- read.csv(url, sep=" ", header = F)
# names <- c("age", "sex", "cp", "restbp", 
#   "chol", "fbs", "restecg", "maxach", "exang", "oldpeak", "slope", "num", 
#   "thal","disease")
# heart<-data.frame(heart)
# colnames(heart) <- names
# heart.dat<-heart
# heart.dat$sex <- factor(heart.dat$sex, labels=c("F", "M"))
# 
# heart.dat$cp <- factor(heart.dat$cp, 
#   labels=c("Typ", "Atyp", "Non-Ang", "Asymp"))
# 
# heart.dat$fbs <- factor(heart.dat$fbs, labels=c("T", "F"))
# 
# heart.dat$restecg <- factor(heart.dat$restecg, 
#   labels=c("Normal", "Abnorm", "Hypertrophy"))
# 
# heart.dat$exang <- factor(heart.dat$exang, labels=c("N", "Y"))
# 
# heart.dat$slope <- factor(heart.dat$slope, 
#   labels=c("Up", "Flat", "Down"))
# 
# heart.dat$thal <- factor(heart.dat$thal, 
#   labels=c("Normal", "Fixed", "Reversible"))
# 
# heart.dat$disease <- factor(heart.dat$disease, labels=c("H", "S"))
```

*NOTE:* the call to the internet may cause problems when knitting the
document to HTML. To solve this you should run the code to create the
heart.dat table. Then save the table to your drive. Create a new code
chunk to load the table from your drive. Comment the code to prevent it
from running during knitting.

```{r}
# to save the data file, only need to do this once then comment out
#save( heart.dat, file= "heart_data.Rdata") 

# to load the data file, need to do this every time, this will load the object heart.dat
load("heart_data.Rdata")
```

### 2. Create an appropriate graph that shows the number of healthy and sick patients by sex.




### 3. Use an appropriate statistical test to determine if there are a similar distreibution of male and female helathy and sick patients.

```{r}
#HINT, here is a function called table() that can create a table of counts of categories. 

table(heart.dat$disease, heart.dat$sex)
```

### 4. Make appropriate graphs for each of the age, resting blood pressure and cholesterol between the helathy and sick patients (one graph for each continuous variable) 



### 5. Make an appropriate statistical test for each of the resting blood pressure and the cholesterl between helathy and sick patients. 



### 6. Males and female often have different values for clincial and physiological variables. Make a graph that compares cholesterol in males and females who are helathy and sick. Describe your observations from the graph and how this compares to the graph in Question 4. 



### 7. Perfrom a statisical test compariing cholesterol levels between sick males and sick females.Discuss this result and compare to the result of question 5. 


### 8. Consider if the variable thal significantly changes the differences in the maximum heart rate achieved. Use an ANOVA and the Tukey post hoc test to statistically test the differences. Make sure to report and interpret the the p-value of the ANOVA and if it passed or failed the Tukey's test.
